NumPy---Array Programming in Python

### 2020-02-04 body.tex
DOC: Address techincia grammar issues
Address technical grammar issues such as - and ,
Substitute some frequent works with synonyms

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community developed, open source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the \emph{de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
\emph{Application Programming Interface} (API) has become widespread as
a language for array programming.

@@

The \emph{shape} of an array determines the number of elements along each axis,
and the number of axes is the array's dimensionality.
For example, a vector of numbers can be stored as a one dimensional array of
shape $N$, while color videos are four dimensional arrays of shape
$(T, M, N, 3)$.

\emph{Strides} are necessary to interpret computer memory, which stores elements
linearly, as multidimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row, column to column, and so forth.
Consider, for example, a 2-D array of floating point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy is able to
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to directly access NumPy array data in memory.

@@

\emph{Indexing} an array returns single elements, subarrays, or elements that satisfy
a specific condition (Fig.~\ref{fig:array-concepts}b).
Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}c).
Wherever possible, indexing that retrieves a subarray returns a {\em view} on
the original array, such that data is shared between the two arrays.
This provides a powerful way to operate on subsets of array data, while
limiting memory usage.

@@

When performing a vectorized operation (such as addition) on two arrays with
the same shape, it is clear what should happen.
Through \emph{broadcasting}, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array,
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied when an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib together with
an advanced interactive environment like IPython \cite{perez2007ipython},
or Jupyter \cite{Kluyver:2016aa} provides a solid foundation for array
programming in Python.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder.
This rich and productive environment has made Python popular
for scientific research.

@@

Many research groups have designed large,
complex scientific libraries, which add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling mathematical optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted usage of
scientific Python.  Examples of its impactful application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  In many fields these tools have become \emph{the
primary} software environment.  NumPy and its ecosystem are commonly
taught in university courses, bootcamps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

NumPy provides in-memory, multidimensional, homogeneously typed
arrays on CPUs.  It runs on machines
ranging from embedded devices to the world's largest supercomputers,
with performance approaching that of compiled languages.  For the
largest part of its existence, NumPy covered the vast majority of
array computation use cases.

%For algorithms that do not fit the array programming approach,
%or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
%that transparently accelerate bottlenecks.
%These tools allow programmers to express concepts using high level array programming without sacrificing efficiency.
%In many cases, fast algorithms already are implemented by libraries in the scientific Python ecosystem; using these primitives, efficient codes can be written in pure %Python.

@@

In recent years, there has been a rapid growth in the development of
specialized hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Routinely, data sets are now larger than a
single machine's memory, and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is currently unable to
utilize such specialized hardware and storage directly.  However, both
distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
to the \emph{paradigm} of array programming: a gap therefore existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Theano and Caffe---have
already ceased development. And each time that a user
decides to try a new technology, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

%CuPy, Sparse and JAX the most prominent examples---while others have
%NumPy-inspired semantics but, often due to their development history, deviate in places.
%Wherever there are such deviations, we often observe users gently requesting,
%and developers evolving the software step by step toward the NumPy API.

@@

Ideally, operating on specialized arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.
To support array operations between external array objects, NumPy
therefore added the capability to act as a central coordination
mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).

@@

  \caption{\textbf{NumPy API and Array Protocol.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{mean} function is called on a a Dask
     array.  The call succeeds by dispatching to the appropriate library implementation
     (i.e., Dask in this case) and results in a new Dask array.  Compare this
     code to the example code in Fig.~\ref{fig:array-concepts}g.
  }\label{fig:array-protocol}

@@

Thanks to these developments users are now capable, for example, of scaling
their computation to distributed systems with Dask. The protocols
\textit{compose} well, so it is even possible to take NumPy code and scale it
to multi-GPU systems with minimal code changes through Dask arrays containing
multiple CuPy arrays \cite{entschev2019}.

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and e.g.,
% https://github.com/pytorch/pytorch/issues/22402

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
individuals, held a powerful attraction for many early contributors.

@@

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
The reasons for its popularity are varied and complex, but include the fact that
NumPy (a) is built on top of Python,
(b) has a culture of collaboration and good software development practices, and
(c) provides a cutting-edge array programming environment for scientific data analysis.

\emph{Python} is an open source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as a primary learning language in universities
has further popularized NumPy as a tool of choice for modern data science.

@@

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.

NumPy began as an attempt to add an array object to Python and became
the foundation of a rich ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
rich data types such as physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich datatypes
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developements...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.

NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
scientific Python ecosystem, but has become the standard API for tensor
computation and a central coordinating mechanism between array types and
technologies in Python.

>>>>

Today, NumPy underpins almost every Python library that does scientific or
numerical computation including SciPy \cite{virtanen2019scipy},
Matplotlib \cite{hunter2007matplotlib}, pandas \cite{mckinney-proc-scipy-2010},
scikit-learn \cite{pedregosa2011scikit}, and
scikit-image \cite{vanderwalt2014scikit}.
It is a community-developed, open-source library, which provides a
multidimensional Python array object along with array-aware functions
that operate on it.
Because of its inherent simplicity, the NumPy array is
the \emph{de facto} exchange format for array data in Python.
The library has such widespread adoption that not only the array object but also its
\emph{Application Programming Interface} (API) has become widespread as
a language for array programming.

@@

The \emph{shape} of an array determines the number of elements along each axis,
and the number of axes is the array's dimensionality.
For example, a vector of numbers can be stored as a one-dimensional array of
shape $N$, while color videos are four-dimensional arrays of shape
$(T, M, N, 3)$.

\emph{Strides} are necessary to interpret computer memory, which stores elements
linearly, as multidimensional arrays.
It describes the number of bytes to move forward in memory to jump from row to
row or column to column.
Consider, for example, a 2-D array of floating-point numbers with shape
$(4, 3)$, where each element occupies 8 bytes in memory.
To move between consecutive columns we need to jump forward 8 bytes in memory,
and to access the next row $3 \times 8 = 24$ bytes.
The strides of that array are therefore $(24, 8)$.  NumPy is able to
store arrays in either C or Fortran memory order, iterating
first over either rows or columns.  This allows external libraries
written in those languages to access NumPy array data in memory directly.

@@

\emph{Indexing} an array returns single elements, subarrays, or elements that satisfy
a specific condition (Fig.~\ref{fig:array-concepts}b).
Arrays can even be indexed using other arrays (Fig.~\ref{fig:array-concepts}c).
Wherever possible, indexing that retrieves a subarray returns a {\em view} on
the original array, such that data is shared between the two arrays.
This provides a powerful way to operate on subsets of array data while
limiting memory usage.

@@

When performing a vectorized operation (such as addition) on two arrays with
the same shape, it is clear what should happen.
Through \emph{broadcasting}, NumPy allows the dimensions to differ, while
still producing results that appeal to intuition.
A trivial example is the addition of a scalar value to an array, but it also
generalizes to more complex examples such as scaling each column of an array
or generating a grid of coordinates.
In broadcasting, one or both arrays are virtually duplicated (that is, without
copying any data in memory), so that the shapes of the operands match
(Fig.~\ref{fig:array-concepts}d).
Broadcasting is also applied when an array is indexed using arrays of
indices (Fig.~\ref{fig:array-concepts}c).

@@

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and use.
SciPy provides fundamental algorithms for scientific computing,
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib, together with
an advanced interactive environment like IPython \cite{perez2007ipython},
or Jupyter \cite{Kluyver:2016aa}, provides a solid foundation for array
programming in Python.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,sunpy2015,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, and provides array testing infrastructure
and build support for integrating with Fortran code and using specialized
compilers and computing platforms.

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs or
notebooks containing both computation and narrative.
Scientific computing beyond exploratory work is often done in a text editor
or an integrated development environment (IDEs) such as Spyder.
This rich and productive environment has made Python popular
for scientific research.

@@

Many research groups have designed large,
complex scientific libraries, which add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling mathematical optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for the analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

The rapid recent growth of data science, machine learning, and
artificial intelligence has further and dramatically boosted the usage of
scientific Python.  Examples of its impactful application, such as the
above, now exist in almost every discipline in the natural and social
sciences.  These tools have become \emph{the
primary} software environment in many fields.  NumPy and its ecosystem are commonly
taught in university courses, boot camps, and summer schools, and are
at the focus of community conferences and workshops worldwide.

@@

NumPy provides in-memory, multidimensional, homogeneously typed
arrays on CPUs.  It runs on machines
ranging from embedded devices to the world's largest supercomputers,
with performance approaching that of compiled languages.  For the
most its existence, NumPy addressed the vast majority of
array computation use cases.

%For algorithms that do not fit the array programming approach,
%or require even faster implementations, a variety of tools exist (e.g., Cython, Numba, Pythran)
%that transparently accelerate bottlenecks.
%These tools allow programmers to express concepts using high-level array programming without sacrificing efficiency.
%In many cases, fast algorithms are already implemented by libraries in the scientific Python ecosystem;
%these primitives simplify writing code in pure Python.

@@

In recent years, there has been a rapid growth in the development of
specialized hardware, including graphics processing
units (GPUs), tensor processing units (TPUs), and field-programmable
gate arrays (FPGAs).  Data sets routinely exceed the memory capacity of a single machine and may be stored on multiple machines or in
the cloud.  Due to its in-memory data model, NumPy is currently unable to
utilize such specialized hardware and storage directly.  However, both
distributed data and the parallel execution of GPUs, TPUs, and FPGAs map well
to the \emph{paradigm} of array programming: a gap, therefore, existed between
available modern hardware architectures and the tools necessary to
leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to the ecosystem)
The community's efforts to fill this gap led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition, there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by xarray \cite{hoyer2017xarray}.

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Theano and Caffe---have
already ceased development. And each time that a user
decides to try new technology, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

%CuPy, Sparse, and JAX the most prominent examples---while others have
%NumPy-inspired semantics but, often due to their development history, deviate in places.
%Wherever there are such deviations, we often observe users gently requesting,
%and developers evolving the software step by step toward the NumPy API.

@@

Ideally, operating on specialized arrays using NumPy functions or semantics would
simply work, so that users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, or distributed arrays, as appropriate.
To support array operations between external array objects, NumPy,
therefore, added the capability to act as a central coordination
mechanism with a well-specified API (Fig.~\ref{fig:ecosystem}).

@@

  \caption{\textbf{NumPy API and Array Protocol.} Through its array protocol, the NumPy API
     can operate on various types of array objects, including those written outside
     of NumPy. In this example, NumPy's \code{mean} function is called on a Dask
     array.  The call succeeds by dispatching to the appropriate library implementation
     (i.e., Dask in this case) and results in a new Dask array.  Compare this
     code to the example code in Fig.~\ref{fig:array-concepts}g.
  }\label{fig:array-protocol}

@@

Thanks to these developments, users are now capable, for example, of scaling
their computation to distributed systems with Dask. The protocols
\textit{compose} well, so it is even possible to take NumPy code and scale it
to multi-GPU systems with minimal code changes through Dask arrays containing
multiple CuPy arrays \cite{entschev2019}.

@@

% should we hint at some of the community interaction part
% they have helped us develop the neps and, e.g.
% https://github.com/pytorch/pytorch/issues/22402

@@

NumPy was initially developed by students, faculty, and researchers to
provide an advanced, open-source array programming library for Python,
which was free to use and unencumbered by license servers, dongles, and the like.
There was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
individuals, held a powerful attraction for many early contributors.

@@

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the broader scientific Python ecosystem and
heralded the current era of widespread use of Python for scientific computing.
The reasons for its popularity are varied and complex but include the fact that
NumPy (a) is built on top of Python,
(b) has a culture of collaboration and good software development practices, and
(c) provides a cutting-edge array programming environment for scientific data analysis.

\emph{Python} is an open-source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as a primary learning language in universities
has further popularized NumPy as a tool of choice for modern data science.

@@

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs an extensive battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

NumPy combines the expressive power of \emph{array programming}, 
the performance of C, and
the readability, usability, and versatility of Python in a mature,
well-tested, well-documented, and community-developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.

NumPy began as an attempt to add an array object to Python and became
the foundation of a vibrant ecosystem of tools.
Now, a large amount of scientific work depends on NumPy being correct,
fast, and stable.
Work on NumPy continues, so that it may meet the needs of its users.  For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
rich data types such as physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich data types
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developments...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.

NumPy's role in the ecosystem continues to evolve. It is no longer just the foundational array library underlying the
scientific Python ecosystem but has become the standard API for tensor
computation and a central coordinating mechanism between array types and
technologies in Python.
